Topics in Algorithms

### 2018-01-10 project/paper.tex

\usepackage[margin=1in]{geometry}  % set the margins to 1in on all sides
\usepackage{graphicx}              % to include figures
\usepackage{amsmath,bm}            % great math stuff
\usepackage{amsfonts}              % for blackboard bold, etc
\usepackage{amsthm}                % better theorem environments
\usepackage{listings}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage[linesnumbered,ruled,vlined,algosection]{algorithm2e}
\usepackage{parskip} 			   % no paragraph indentation
\usepackage{subcaption}
\usepackage{pgfplots}
\usetikzlibrary{arrows,automata}

@@

\begin{abstract}
  This comprises our project covering Approximation Distance Oracles with Improved Query Time for the Topics in Algorithms course offered at The Department of Computer Science, Uni. Copenhagen.
\end{abstract}

@@

\section{Overview}
% Assumptions about the reader

\begin{figure}
\begin{subfigure}[b]{\textwidth}
\begin{tikzpicture}
\filldraw (0,0) circle (2pt) node[align=left,   below] {$u$};
\filldraw (10,0) circle (2pt) node[align=right,  below] {$v$};
\filldraw[fill=gray] (4,3) circle (2pt) node[align=left, above] {$p_j(u)$};
\filldraw[fill=white] (8,3.45) circle (2pt) node[align=left, above] {$p_{j+1}(v)$};
\coordinate (u) at (0,0) ;
\coordinate (v) at (10,0);
\coordinate (pu) at (4,3);
\coordinate (pv) at (8, 3.45);
\draw[dashed] (v) circle [radius=4];
\draw (u) -- (pu) node[pos=0.5, above, sloped] {$d_G(u,p_j(u))$};
\draw (u) -- (v) node[pos=0.5, below, sloped] {$d_G(u,v))$};
\draw (v) -- (pv) node[pos=0.5, below, sloped] {$d_G(v, p_{j+1}(v)$};
\draw[dotted] (v) -- (pu) node[pos=0.75, below, sloped] {$d_G(v, p_{j}(u)$};
\end{tikzpicture}
\caption{$(j)$th iteration\label{j1}}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\begin{tikzpicture}
\filldraw (0,0) circle (2pt) node[align=left,   below] {$v$};
\filldraw (10,0) circle (2pt) node[align=right,  below] {$u$};
\filldraw[fill={rgb:black,1;white,3}] (5,4.9) circle (2pt) node[align=left, above] {$p_{j+2}(v)$};
\filldraw[fill=white] (8,3.45) circle (2pt) node[align=left, above] {$p_{j+1}(u)$};
\coordinate (v) at (0,0) ;
\coordinate (u) at (10,0);
\coordinate (pv) at (5,4.9);
\coordinate (pu) at (8, 3.45);
\begin{scope}
    \clip (0,6) rectangle (7,-1);
    \draw[dashed] (v) circle [radius=7];
\end{scope}

\draw (v) -- (pv) node[pos=0.5, above, sloped] {$d_G(v,p_{j+2}(v))$};
\draw (v) -- (u) node[pos=0.5, below, sloped] {$d_G(v,u))$};
\draw (u) -- (pu) node[pos=0.5, below, sloped] {$d_G(u, p_{j+1}(u)$};
\draw[dotted] (v) -- (pu) node[pos=0.5, below, sloped] {$d_G(v, p_{j+1}(u)$};
\end{tikzpicture}
\caption{$(j+1)$th iteration.\label{j2}}
\end{subfigure}
\caption{Two consecutive iterations of the while loop in $dist_k$. Both sampled points of $u$, $p_j(u)$ in Fig. \ref{j1} and $p_{j+1}(u)$ in Fig. \ref{j2}, are not in $B_v$ and therefore $j$ is not $(u,v)$-terminal.}
\end{figure}
% Reference til Mikkel/Zwick paper - hans query time og 'vores' forbedrede query time - forbedring af konstruktion. Forbedring i querytime, uden at det påvirker space/konstruktionstid.

@@

Given an undirected, positively weighted graph $G$ with $m$ edges, $n$ vertices and given $k \in \mathbb{Z}$, $k \geq 2$, we show that a $2k-1$ approximate distance oracle for $G$ of size $O(kn^{1+1/k})$ and with $O(\lg k)$ query time can be constructed in $O( \min \left\{ kmn^{1/k}, \sqrt{k}m+kn^{1+c/\sqrt{k}} \right\})$ time for some constant $c$.
% Definitioner
\begin{itemize}
\item $d_{G}(u,v)$: the shortest path distance between $u$ and $v$.
\item $A_i = \left\{ a \in A_{i-1} \; | \; a \text{ sampled with prob. } n^{-1/k} \text{ uniformly, i.i.d.} \right\}$
\item $V = A_0 \supseteq A_1 \supseteq \hdots \supseteq A_k = \emptyset$
\item $p_i(u) = \min_{v \in A_i} \left\{ d_{G}(u,v) \right\}$
\item $B_u=\bigcup_{i=0}^{k-1}\left\{ v \in A_i - A_{i+1} \; | \; d_{G}(u,v) < d_{G}(u, p_{i+1}(u)) \right\}$
\item $d_G(u,p_k(u))=\infty$
\end{itemize}

\begin{lem}
Let $u,v \in V$ be distinct vertices $0 \leq i < k-1$. If $p_i(v) \not\in B_{u}$ then $d_G(u, p_{i+1}(u)) \leq d_G(u, p_i(v))$. Furthermore $A_{k-1} \subset B_u$ and $p_{k-1}(v) \in B_u$. \label{lem:onedeltaperiteration}
\end{lem}

@@

\paragraph{Proof} If $p_j(u) \in B_v$ or $p_j+1(v) \in B_u$, then $j$ is $(u,v)$-feasible, and so is $i_1, \hdots, j$. Assume that $p_j(u) \not\in B_v$ and $p_j+1(v) \not\in B_u$. Then, by Lemma \ref{lem:lemma1}, $d_G(v,p_{j+1}(v)) \leq d_G(v,p_{j}(u))$ and $d_G(u,p_{j+2}(u)) \leq d_G(u,p_{j+1(v)})$. Applying the triangle inequality twice, we get 
\begin{align*}
d_G(u,p_{j+2}(u) &\leq d_G(u,p_{j+1}(v)) \\
&\leq d_G(u,v) + d_G(v, p_{j+1}(v)) \\
&\leq d_G(u,v) + d_G(v, p_{j}(u)) \\
&\leq 2d_G(u,v) + d_G(u, p_{j}(u))
\end{align*}
so $\delta_j(u) = d_G(u,p_{j+2}(u)) - d_G(u,p_j(u)) \leq 2d_G(u,v)$. Let $\mathcal{I}'$ be the set of even indices $i_1, i_1+2, i_1+4, \hdots, i-2$. Since $i_1, \hdots, i_2$ is $(u,v)$-feasible, $d_G(u,p_{i_1}(u)) \leq i_1 \cdot d_G(u,v)$. By the choice of $j$,
\begin{align*}
d_G(u,p_{i}(u)) &= d_G(u,p_{i_1}(u)) + \sum_{j' \in \mathcal{I'}}\delta_{j'}(u) \\
&\leq d_G(u,p_{i_1}(u)) + \left| \mathcal{I'} \right| \max_{j' \in \mathcal{I'}}\delta_{j'}(u) \\
&= d_G(u,p_{i_1}(u)) + \frac{i-i_1}{2} \delta_{j}(u) \\
&\leq i_1 \cdot d_G(u,v) + (i-i_1)d_G(u,v) \\
&= i \cdot d_G(u,v)
\end{align*}
Hence, since $i_1, \hdots, i_2$ is $(u,v)$-feasible, so is $i, \hdots, i_2$.
\end{lem}

\begin{figure}[H]
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=3cm},
  level 2/.style={sibling distance=1.5cm}]
  \node[color=red] {$I'$}
    child[color=red] {node {$[0, ..., n/2]$}
    child[color=black,thick] {node[color=cyan] {$I^j$}}
      child[dashed] {node {$p(l_1)$}
      	child[solid, left=0.1cm, color=black] {node[color=cyan] {$l_1=[i_1, i_1+2]$}}
      	child[solid,color=black] {node[color=cyan] {$[i_1+2,i_1+4]$}}
      }
    }
    child[right=0.1cm, color=red] {node {$[n/2+2,...,n]$}
    child[color=black] {node[color=cyan] {$I^k$} 
    						child[dotted, color=black] {node {...}}
    						child[dotted, color=black,left=0.1cm] {node {...}}
    					}
      child[dashed] {{node {$p(l_2)$}
       	child[solid,color=black] {node[color=cyan] {$[i_2-4,i_2-2]$}}
      	child[right=1cm,solid,color=black] {node[color=cyan] {$l_2=[i_2-2, i_2]$}}
    }}};

@@

>>>>

\usepackage[margin=1in]{geometry}  % set the margins to 1in on all sides
\usepackage{graphicx}              % to include figures
\usepackage{amsmath,bm}            % great math stuff
\usepackage{amsfonts}              % for blackboard bold, etc
\usepackage{amsthm}                % better theorem environments
\usepackage{breqn}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{tikzscale}
\usepackage{pgfplots}
\usepackage[linesnumbered,ruled,vlined,algosection]{algorithm2e}
\usepackage{parskip} 			   % no paragraph indentation
\usetikzlibrary{arrows,automata,decorations.pathreplacing,math,arrows,automata}
\usepackage{subcaption}

@@

\begin{abstract}
  This comprises our project covering the first half of the paper "Approximation Distance Oracles with Improved Query Time" by Christian Wulff-Nilsen \cite{Wulff2013}. This project attempts to elaborate and provide intuition on the $\lg k$-query time improvement by providing explanations and visualisations of proofs.
\end{abstract}

@@

\section{Overview}
Given an undirected graph $G$ with non-negative edge weights, $m$ edges, $n$ vertices and some $k \in \mathbb{Z} \geq 2$, Thorup and Zwick showed that a $(2k-1)$-approximate distance oracle for $G$ of size $O(kn^{1+1/k})$ and $O(k)$ query time can be constructed in $kmn^{1/k}$ expected time \cite{Thorup2005}. \cite{Wulff2013} improves this query time to $\lg k$, while preserving stretch, space and preprocessing time. % Specifically $O( \min \left\{ kmn^{1/k}, \sqrt{k}m+kn^{1+c/\sqrt{k}} \right\})$ time for some constant $c$.

The reduction in query time is accomplished by performing binary search on the bunch structures described in \cite{Thorup2005} by uniquely determining at which level the bunch 
% Assumptions about the reader

% Reference til Mikkel/Zwick paper - hans query time og 'vores' forbedrede query time - forbedring af konstruktion. Forbedring i querytime, uden at det påvirker space/konstruktionstid.

@@

\section{Definitions}
The following definitions are taken directly from \cite{Thorup2005} and \cite{Wulff2013} and are used throughout this project.
\begin{itemize}
\item $d_{G}(u,v)$: the shortest path distance between $u$ and $v$.
\item $A_i = \left\{ a \in A_{i-1} \; | \; a \text{ sampled with prob. } n^{-1/k} \text{ uniformly, i.i.d.} \right\}$
\item $V = A_0 \supseteq A_1 \supseteq \hdots \supseteq A_k = \emptyset$
\item $p_i(u) = \min_{v \in A_i} \left\{ d_{G}(u,v) \right\}$
\item $B_u=\bigcup_{i=0}^{k-1}\left\{ v \in A_i - A_{i+1} \; | \; d_{G}(u,v) < d_{G}(u, p_{i+1}(u)) \right\}$, the bunch of $u$.
\item $d_G(u,p_k(u))=\infty$
\end{itemize}
Thorup and Zwick showed that
\\
\begin{lem}
Let $u,v \in V$ be distinct vertices $0 \leq i < k-1$. If $p_i(v) \not\in B_{u}$ then $d_G(u, p_{i+1}(u)) \leq d_G(u, p_i(v))$. Furthermore $A_{k-1} \subset B_u$ and $p_{k-1}(v) \in B_u$. \label{lem:onedeltaperiteration}
\end{lem}

@@

\paragraph{Proof} If $p_j(u) \in B_v$ or $p_j+1(v) \in B_u$, then $j$ is $(u,v)$-feasible, and so is $i_1, \hdots, j$. Assume that $p_j(u) \not\in B_v$ and $p_j+1(v) \not\in B_u$. Then, by Lemma \ref{lem:onedeltaperiteration}, $d_G(v,p_{j+1}(v)) \leq d_G(v,p_{j}(u))$ and $d_G(u,p_{j+2}(u)) \leq d_G(u,p_{j+1(v)})$. Applying the triangle inequality twice, cfr. Fig. \ref{fig:twoiterations}, we get 
\begin{align*}
d_G(u,p_{j+2}(u) &\leq d_G(u,p_{j+1}(v)) &(\text{Lemma \ref{lem:onedeltaperiteration}}) \\
&\leq d_G(u,v) + d_G(v, p_{j+1}(v)) &(\text{Triangle Inequality})\\
&\leq d_G(u,v) + d_G(v, p_{j}(u)) &(\text{Lemma \ref{lem:onedeltaperiteration}})\\
&\leq 2d_G(u,v) + d_G(u, p_{j}(u))&(\text{Triangle Inequality})
\end{align*}
so $\delta_j(u) = d_G(u,p_{j+2}(u)) - d_G(u,p_j(u)) \leq 2d_G(u,v)$. Let $\mathcal{I}'$ be the set of even indices $i_1, i_1+2, i_1+4, \hdots, i-2$. Since $i_1, \hdots, i_2$ is $(u,v)$-feasible, $d_G(u,p_{i_1}(u)) \leq i_1 \cdot d_G(u,v)$. By the choice of $j$,
\begin{align*}
d_G(u,p_{i}(u)) &= d_G(u,p_{i_1}(u)) + \sum_{j' \in \mathcal{I'}}\delta_{j'}(u) \\
&\leq d_G(u,p_{i_1}(u)) + \left| \mathcal{I'} \right| \max_{j' \in \mathcal{I'}}\delta_{j'}(u) \\
&= d_G(u,p_{i_1}(u)) + \frac{i-i_1}{2} \delta_{j}(u) \\
&\leq i_1 \cdot d_G(u,v) + (i-i_1)d_G(u,v) \\
&= i \cdot d_G(u,v)
\end{align*}
Hence, since $i_1, \hdots, i_2$ is $(u,v)$-feasible, so is $i, \hdots, i_2$.  \label{lem:uvfeasibility}
\end{lem}

\begin{figure}[!hbt]
  \centering
  \subcaptionbox{$(j)$th iteration\label{fig:j1}}{
    \includegraphics[width=.45\textwidth]{include/tikz/jth_iteration.tikz}%
  }
  \hfill
  \subcaptionbox{$(j+1)$th iteration\label{fig:j2}}{
    \includegraphics[width=.45\textwidth]{include/tikz/j1th_iteration.tikz}%
  }
  \caption{Two consecutive iterations of the while loop in $dist_k$. Both sampled points of $u$, $p_j(u)$ in Fig. \ref{fig:j1} and $p_{j+1}(u)$ in Fig. \ref{fig:j2}, are not in $B_v$ and therefore $j$ is not $(u,v)$-terminal. An upper bound of $2d_{G}(u,v)+d_G(u,p_j(u))$ is derived on $d_G(u,p_{j+2}(u))$ in Lemma \ref{lem:uvfeasibility} by applying Lemma \ref{lem:onedeltaperiteration} and the triangle inequality twice. \label{fig:twoiterations}}
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=1\textwidth]{include/tikz/distance_real_line.tikz}%
\caption{Visualization of the distance $d_G(u,p_j(u))$ and $\delta_j(u)$. The green lines correspond to the circles defined by $u$ and the radii $d_G(p_l(u))$ for $l \in \left\{ i_1, j, j+2, i-2, i, i_2 \right\}$.}
\end{figure}

\begin{figure}[h]
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=3cm},
  level 2/.style={sibling distance=1.5cm}]
  \node[color=red] {$I'$}
    child[color=red] {node {$[0, ..., n/2]$}
    child[color=black,thick] {node[color=cyan] {$I^j$}}
      child[dashed] {node {$p(l_1)$}
      	child[solid, left=0.1cm, color=black] {node[color=cyan] {$l_1=[i_1, i_1+2]$}}
      	child[solid,color=black] {node[color=cyan] {$[i_1+2,i_1+4]$}}
      }
    }
    child[right=0.1cm, color=red] {node {$[n/2+2,...,n]$}
    child[color=black] {node[color=cyan] {$I^k$} 
    						child[dotted, color=black] {node {...}}
    						child[dotted, color=black,left=0.1cm] {node {...}}
    					}
      child[dashed] {{node {$p(l_2)$}
       	child[solid,color=black] {node[color=cyan] {$[i_2-4,i_2-2]$}}
      	child[right=1cm,solid,color=black] {node[color=cyan] {$l_2=[i_2-2, i_2]$}}
    }}};

@@

\bibliographystyle{plain}
\bibliography{bibliography}
